---
title: "Analysis of FEV Data Set"
author: "Michael Williams"
date: "May 28, 2015"
output: pdf_document
---

\section{Overview}

The accompanying R code gives a demonstration of data manipulation, graphics, and statistical procedures. The outputs contain data displays (histograms, boxplots, scatterplots) and statistical analysis (categorical data analysis, tTest for comparison of means, linear regression).  

The data correspond to the following variables and descriptions.  

* **id** for *Identification Number*  
*  **age** for *Age* (in years)  
*  **fev** for *Forced Expiratory Volume* (in liters)  
*  **height** for *Height* (in inches)  
*  **sex** for *Sex* ('male' or 'female')  
*  **smoke** for *Smoking Status* ('non-current smoker' or 'current smoker')  

The data was obtained from B. Rosner's book *Introduction to Biostatistics*.

\section{Exploratory Analysis of FEV}

In this section, we explore the data by producing summary statistics and insightful graphics. For the graphics, the ggplot2 package will be used.

```{r}
library(ggplot2)
```

Load the FEV data set via 
```{r}
load(url('http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/FEV.sav')) 
```

The first six rows of FEV.
```{r}
head(FEV) 
```

Summary statistics for all variables.
```{r}
summary(FEV)
```

Scatter plot of fev by age with smoke indicated.
```{r}
plot1 <- ggplot(FEV, aes(x=age, y=fev, colour=smoke)) + geom_point()
plot1 <- plot1 + labs(x="Age", y="Forced Expiratory Volume", 
                      title="FEV by Age with Indicated Smoking Status")
plot1
```

It will be useful to find out the age of the youngest male smokers.
```{r}
summary(
        subset(
                FEV,
                subset = smoke=="current smoker" & sex=="male",
                select = age
               )
        )
```

We do the same for the female smokers.
```{r}
summary(
        subset(
                FEV,
                subset = smoke=="current smoker" & sex=="female",
                select = age
               )
        )
```

Create a contingency table for smoke by sex.
```{r}
cats <- subset(FEV, select = c(sex,smoke))
tcats <- table(cats)
tcats
```

The code `prop.table(tcats, margin=1)` gives proporitons by sex.
```{r}
prop.table(tcats, margin=1)
```

Use a chi-square test for proportions without Yates correction.
```{r}
chisq.test(tcats, correct=FALSE)
```

Boxplot of fev by smoke and sex.
```{r}
plot2 <- ggplot(FEV, aes(smoke, fev)) + geom_boxplot(aes(fill = sex)) 
plot2 <- plot2 + labs(title="Boxplots of fev by smoke for each gender")
plot2
```

Finally, we give a linear regression model of the response variable fev and predictor variables age, height, sex, and smoke.

```{r}
fit <- lm(fev ~ age + height + sex + smoke, data=FEV)
summary(fit)
```

\section{FEV Restricted to Teenagers}

In this section, we restrict the FEV data set to teenagers, that is, those subjects age 13 to 19. This restricted data set is called FEV2.

```{r}
FEV2 <- subset(FEV, subset = age>= 13)
```

Scatterplot of fev by age with smoke coding.
```{r}
plot3 <- ggplot(FEV2, aes(x=age, y=fev, colour=smoke)) + geom_point()
plot3 <- plot3 + labs(x="Age", y="Forced Expiratory Volume", 
                      title="FEV by Age with Indicated Smoking Status\nAge 13+")
plot3
```

A table for smoke status by sex.
```{r}
cats2 <- subset(FEV2, select = c(sex,smoke))
tcats2 <- table(cats2)
tcats2
```

The code `prop.table(tcats2, 1)` gives proporitons by sex.
```{r}
prop.table(tcats2, 1)
```

Use a chi-square test for proportions without Yates correction.
```{r}
chisq.test(tcats2, correct=FALSE)
```

Boxplot of FEV by smoke and sex.
```{r}
plot4 <- ggplot(FEV2, aes(smoke, fev)) + geom_boxplot(aes(fill = sex)) 
plot4 <- plot4 + labs(title="Boxplots of FEV by smoke and sex\nAge 13+")
plot4
```

Histogram of fev for males.
```{r}
plot5 <- ggplot(subset(FEV2, subset = sex == "male"), 
                aes(fev)) + geom_histogram()
plot5 <- plot5 + labs(x="Forced Expiratory Volume", 
                      title="fev for Males\nAge 13+")
plot5
```

Histogram of fev for females.
```{r}
plot6 <- ggplot(subset(FEV2, subset = sex == "female"),
                aes(fev)) + geom_histogram()
plot6 <- plot6 + labs(x="Forced Expiratory Volume", 
                      title="fev for Females\nAge 13+")
plot6
```

t-test for fev by sex.
```{r}
t.test(fev ~ sex, data = FEV2)
```

Split-up the data set FEV2 by sex
```{r}
FEV2m <- subset(FEV2, subset = sex == "male")
FEV2f <- subset(FEV2, subset = sex == "female")
```

t-test for fev by smoke.
```{r}
t.test(fev ~ smoke, data = FEV2m)
```

t-test for fev by smoke.
```{r}
t.test(fev ~ smoke, data = FEV2f)
```

scatterplot of fev by age with smoke coding for males.
```{r}
plot7 <- ggplot(FEV2m, aes(x=age, y=fev, colour=smoke)) + geom_point()
plot7 <- plot7 + labs(x="Age", y="Forced Expiratory Volume", 
                      title="FEV by Age with Indicated Smoking Status\nMales Age 13+")
plot7
```

scatterplot of fev by age with smoke coding for females.
```{r}
plot8 <- ggplot(FEV2f, aes(x=age, y=fev, colour=smoke)) + geom_point()
plot8 <- plot8 + labs(x="Age", y="Forced Expiratory Volume", 
                      title="FEV by Age with Indicated Smoking Status\nFemales Age 13+")
plot8
```